The Design and Implementation of the Synchronous Language \CEU

### 2016-10-04 paper.tex
(>) Figura "terra"
(>) \emph all quotes

Another distinction regarding event handling in comparison to \CEU is that 
Esterel supports same-cycle bi-directional 
communication~\cite{esterel.compiling}, i.e., two threads can react to one 
another during the same cycle due to mutual signal dependencies
(a.k.a. \emph{instantaneous dialogue}~\cite{rp.book}).
%
\CEU takes a different approach, posing a tradeoff that an \code{await} is only
valid for the next reaction, i.e., if an \code{await} and \code{emit} occur 
simultaneously in parallel trails, the \code{await} does not awake.
%
This approach avoids any form of graph construction or data dependency analysis
as required in Esterel, e.g.:
``set or reset of a signal must always precede any test of this signal (signal
coherence law and constructivity''~\cite{esterel.saxort}.
%
Also, these \emph{delayed awaits} avoid corner cases of instantaneous
termination and re-execution of statements in the same reaction (known as
\emph{schizophrenic statements}~\cite{esterel.constructive}).

@@

Esterel is only deterministic with respect to external behavior: ``the same 
sequence of inputs always produces the same sequence of 
outputs''~\cite{esterel.primer}.
%
However, the execution order for operations within a reaction is 
non-deterministic: ``if there is no control dependency, as in
\code{(call~f1()~||~call~f2())},
the order is unspecified and it would be an error to rely on 
it''~\cite{esterel.primer}.
%
For this reason, Esterel, does not support shared-memory concurrency:
\emph{``if a variable is written by some thread, then it can neither be read
nor be written by concurrent threads''}~\cite{esterel.primer}.
%
A number of Esterel-inspired synchronous languages enforce an arbitrary 
execution order for statements in multiple lines of execution to achieve 
intra-reaction determinism
(%
\emph{Reactive C}~\cite{rp.rc},                 % 1991
\emph{Protothreads}~\cite{wsn.protothreads},    % 2006
\emph{SOL}~\cite{wsn.sol},                      % 2007
\emph{SC}~\cite{rp.synchc},                     % 2009
and
\emph{PRET-C}~\cite{rp.pretc}%                  % 2010
).
%, e.g., based on lexical order or manual priority assignment.
%
\CEU also takes the deterministic approach and, when multiple trails are active 
during the same reaction, they are scheduled in lexical order, i.e., in the 
order they appear in the program source code.

@@

In future work, we plan to perform an equivalent throughout analysis in
comparison to industrial-quality code bases in Esterel.
%
An Esterel implementation that compiles to sequential C code with similar
techniques also generates ``code less than $10\%$ larger the original
manually-sequenced code'' for an industrial-size
application~\cite{esterel.efficient}.
%
%Unfortunately, most real applications in Esterel seem to be closed source.
%In future work, we plan to cooperate with groups that have more experience 
%and access to real code bases in Esterel for a comprehensive comparison with 
%\CEU.
% TODO-no:
%Still, considering \C as an ``absolute benchmark'', we consider the small 
%overhead .  solid results

@@

TODO:
\begin{figure}[!htb]
 \centering
 %\includegraphics[width=0.5\columnwidth]{terra}
\caption{Terra programming system basic elements.}
 \label{fig.terra}
\end{figure}

@@

In \CEU, internal events support stack-based micro reactions within external
reactions, providing more fine-grained control for intra-reaction execution.
%
Some variants of Statecharts~\cite{statecharts} also distinguish internal from
external events~\cite{statecharts.variants}.
%Some variants of Statecharts support local events which imply handling intra
%reactions like \CEU.
In Statemate, ``reactions to external and internal events (...) can be sensed only
after completion of the step''~\cite{statecharts.statemate}, implying queue-based
execution.
In Stateflow, ``the receiving state (of the event) acts here as a
function''~\cite{statecharts.stateflow}, which is similar to \CEU's stack-based
execution.
To avoid recursion, \CEU adopts delayed awaits, while in Statemate, ``loops in
the broadcasting of events (between states) are forbidden''.

Like \CEU, many synchronous languages rely on deterministic scheduling to 
preserve intra-reaction determinism (%
\emph{Reactive C}~\cite{rp.rc},                 % 1991
\emph{Protothreads}~\cite{wsn.protothreads},    % 2006
\emph{SOL}~\cite{wsn.sol},                      % 2007
\emph{SC}~\cite{rp.synchc},                     % 2009
and
\emph{PRET-C}~\cite{rp.pretc}).                 % 2010
%
%It is the responsibility of the programmer to specify the execution order for 
%threads, based on either explicit priorities, or source code lexical order 
%(similar to \CEU).
%
\CEU also performs concurrency checks to detect trails that, when reordered,
change the observable behavior of the program, i.e., trails that actually rely
on deterministic scheduling.
%
%These languages have a tick-based notion of time similar to Esterel, which 
%prevents the concurrency checks of \CEU.
\emph{SCCharts}~\cite{sccharts.1} proposes \emph{sequentially constructive
programs} that allow ``shared variables to have multiple values per tick as
long as these values are explicitly ordered by sequential
statements''~\cite{sccharts.2}.
This relaxes Esterel's restriction for \emph{read-after-write} accesses only
(e.g., a signal must be emitted to be present) and permits
\emph{write-after-read} accesses typical in imperative programs (e.g., test and
set).
Although this approach still restricts concurrent accesses, it does not impose
strict deterministic scheduling such as in \CEU.

>>>>

Another distinction regarding event handling in comparison to \CEU is that 
Esterel supports same-cycle bi-directional 
communication~\cite{esterel.compiling}, i.e., two threads can react to one 
another during the same cycle due to mutual signal dependencies
(a.k.a. \emph{instantaneous dialogue}~\cite{rp.book}).
%
\CEU takes a different approach, posing a tradeoff that an \code{await} is only
valid for the next reaction, i.e., if an \code{await} and \code{emit} occur 
simultaneously in parallel trails, the \code{await} does not awake.
%
This approach avoids any form of graph construction or data dependency analysis
as required in Esterel, e.g.:
\emph{``set or reset of a signal must always precede any test of this signal
(signal coherence law and constructivity''}~\cite{esterel.saxort}.
%
Also, these \emph{delayed awaits} avoid corner cases of instantaneous
termination and re-execution of statements in the same reaction (known as
\emph{schizophrenic statements}~\cite{esterel.constructive}).

@@

Esterel is only deterministic with respect to external behavior: \emph{``the
same sequence of inputs always produces the same sequence of 
outputs''}~\cite{esterel.primer}.
%
However, the execution order for operations within a reaction is 
non-deterministic: \emph{``if there is no control dependency, as in
\code{(call~f1()~||~call~f2())},
the order is unspecified and it would be an error to rely on 
it''}~\cite{esterel.primer}.
%
For this reason, Esterel, does not support shared-memory concurrency:
\emph{``if a variable is written by some thread, then it can neither be read
nor be written by concurrent threads''}~\cite{esterel.primer}.
%
A number of Esterel-inspired synchronous languages enforce an arbitrary 
execution order for statements in multiple lines of execution to achieve 
intra-reaction determinism
(%
\emph{Reactive C}~\cite{rp.rc},                 % 1991
\emph{Protothreads}~\cite{wsn.protothreads},    % 2006
\emph{SOL}~\cite{wsn.sol},                      % 2007
\emph{SC}~\cite{rp.synchc},                     % 2009
and
\emph{PRET-C}~\cite{rp.pretc}%                  % 2010
).
%, e.g., based on lexical order or manual priority assignment.
%
\CEU also takes the deterministic approach and, when multiple trails are active 
during the same reaction, they are scheduled in lexical order, i.e., in the 
order they appear in the program source code.

@@

In future work, we plan to perform an equivalent throughout analysis in
comparison to industrial-quality code bases in Esterel.
%
An Esterel implementation that compiles to sequential C code with similar
techniques also generates \emph{``code less than $10\%$ larger the original
manually-sequenced code''} for an industrial-size
application~\cite{esterel.efficient}.
%
%Unfortunately, most real applications in Esterel seem to be closed source.
%In future work, we plan to cooperate with groups that have more experience 
%and access to real code bases in Esterel for a comprehensive comparison with 
%\CEU.
% TODO-no:
%Still, considering \C as an ``absolute benchmark'', we consider the small 
%overhead .  solid results

@@

\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.5\columnwidth]{terra}
    \caption{Terra programming system basic elements.}
    \label{fig.terra}
\end{figure}

@@

In \CEU, internal events support stack-based micro reactions within external
reactions, providing more fine-grained control for intra-reaction execution.
%
Some variants of Statecharts~\cite{statecharts} also distinguish internal from
external events~\cite{statecharts.variants}.
%Some variants of Statecharts support local events which imply handling intra
%reactions like \CEU.
In Statemate, \emph{``reactions to external and internal events (...) can be
sensed only after completion of the step''}~\cite{statecharts.statemate},
implying queue-based execution.
In Stateflow, \emph{``the receiving state (of the event) acts here as a
function''}~\cite{statecharts.stateflow}, which is similar to \CEU's
stack-based execution.
To avoid recursion, \CEU adopts delayed awaits, while in Statemate,
\emph{``loops in the broadcasting of events (between states) are forbidden''}.

Like \CEU, many synchronous languages rely on deterministic scheduling to 
preserve intra-reaction determinism (%
\emph{Reactive C}~\cite{rp.rc},                 % 1991
\emph{Protothreads}~\cite{wsn.protothreads},    % 2006
\emph{SOL}~\cite{wsn.sol},                      % 2007
\emph{SC}~\cite{rp.synchc},                     % 2009
and
\emph{PRET-C}~\cite{rp.pretc}).                 % 2010
%
%It is the responsibility of the programmer to specify the execution order for 
%threads, based on either explicit priorities, or source code lexical order 
%(similar to \CEU).
%
\CEU also performs concurrency checks to detect trails that, when reordered,
change the observable behavior of the program, i.e., trails that actually rely
on deterministic scheduling.
%
%These languages have a tick-based notion of time similar to Esterel, which 
%prevents the concurrency checks of \CEU.
\emph{SCCharts}~\cite{sccharts.1} proposes \emph{sequentially constructive
programs} that allow \emph{``shared variables to have multiple values per tick
as long as these values are explicitly ordered by sequential
statements''}~\cite{sccharts.2}.
This relaxes Esterel's restriction for \emph{read-after-write} accesses only
(e.g., a signal must be emitted to be present) and permits
\emph{write-after-read} accesses typical in imperative programs (e.g., test and
set).
Although this approach still restricts concurrent accesses, it does not impose
strict deterministic scheduling such as in \CEU.
